### 2015-06-30 text/ms.Rmd
Expand bbmsy macros for RTF and Pandoc

What can we do when multiple ecological models suggest different conclusions about population status? Any single model is unlikely to perform best under all scenarios. Ensemble methods, commonly used in fields such as climate science, provide a powerful tool to leverage information across multiple ecological models. Here we evaluate the potential for ensemble methods to improve estimates of fish population status and trajectory. We fit four data-limited models of population and exploitation dynamics and combine their estimates of the mean and slope of biomass at maximum sustainable yield (\bbmsy\\) with four ensemble models: a simple average, a linear model, and two machine-learning methods---random forests and boosted regression trees. We build our ensembles on a simulated dataset of 5760 fish stocks and test our models with cross-validation and a global database of `r ram_stocks_n` stock assessments. We find that ensemble models substantially improve estimates of population status and trajectory. Machine-learning ensembles performed the best: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, correlation improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`. We found similar improvements when predicting the slope of status. When the simulation-trained ensembles were applied to global fish stocks, the best ensembles still improved all dimensions of performance over the next-best individual model. Ensemble models have potential to improve estimates of population status and trajectory across many taxa beyond fish. 
However, ensemble methods must be carefully tested, formed from a diverse set of accurate models, and built on a dataset sufficiently representative of the populations they are applied to.

@@

We fit four individual data-limited models to estimate \bbmsy\\. Three of the
models are mechanistic and based generally on Schaefer biomass dynamics
[@schaefer1954] of the form

@@

* *mPRM* (modified panel regression model) is a modified version of the
  panel-regression model from Costello et al. [@costello2012]. Unlike the other
  models, mPRM is empirical and not mechanistic---it uses the RAM
  Legacy Stock Assessment database to fit a regression model to a series of
  characteristics of the catch time series and stock with stock-assessed
  \bbmsy\\ as the response. The model used in this paper is
  modified from the original in that it condenses the life-history categories
  into three categories to match the simulated dataset, removes the maximum
  catch predictor since the absolute catch in the simulated dataset is
  arbitrary, and does not implement the bias correction needed in Costello et
  al. [@costello2012] for deriving aggregate estimates of mean status across
  multiple stocks.

@@

The individual models we seek to combine with ensemble models provide time
series of \bbmsy\\. Therefore, we can use ensemble models to estimate multiple
properties of these status time series. Here, we focus on two properties: the
mean and slope of \bbmsy\\ in the last 5 years. Together, these quantities
address the recent state and trajectory of status, which are both of management
and conservation interest (e.g. [@mace2008]). To avoid undue influence of the
end points of the time series on the calculated slope, we measured the slope as
the Theil-Sen estimator of median slope [@theil1950].

@@

\noindent 
where $L$ is a generalized loss function, $f$ is some generalized regression
function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY
model etc. (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the
spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $b$
is the mean or slope value the model is fitted to, and $\theta$ is the set of
parameters for the regression function. For all ensemble models of mean
\bbmsy\\---a ratio bounded at zero---we fit the models in log space and
exponentiated the predictions. For the estimates of \bbmsy\\ slope, we fit
ensemble models on the natural untransformed scale.

We calculated the average ensemble models as the geometric mean of the
individual model \bbmsy\\ estimates and the mean of the individual model
\bbmsy\\ slopes. We fit the linear model ensemble with second-order
interactions. We fit two machine learning ensemble models based on regression
trees. Regression trees sequentially determine what value of a predictor best
splits the response data into two 'branches' based on a specified loss
function [@breiman1984]. In random forests, a series of regression trees are
built on a random subset of the data and random subset of the covariates of the
model [@breiman2001]. In generalized boosted models (GBMs), each subsequent
model is fit to the residuals from the previous model; data points that are fit
poorly in a given model are given more weight in the next model [@elith2008].
Random forests and GBMs can provide strong predictive performance and fit
highly non-linear relationships [@elith2008; @hastie2009]. We fit random forest
models with the `randomForest` package [@liaw2002] for \textsf{R} with the
default argument values. We fit boosted regression tree models with the `gbm`
package [@ridgeway2015] for \textsf{R}. Based on cross-validation with the
**caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth
of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other
arguments at their default values.

@@

Predictive performance can be evaluated with metrics that represent a variety
of modelling goals. For continuous response variables such as the mean and
slope of population status, performance metrics often measure some form of
bias, precision, accuracy (a combination of bias and precision), or the ability
to correctly rank or correlate across populations (e.g. [@walther2005]). Here,
we measure proportional error, defined as
$(\hat{b} - b)/b$, where $\hat{b}$ and $b$ represent estimated and 'true'
(or stock-assessed) mean or slope of \bbmsy\\. We calculated median proportional
error to measure bias, median absolute proportional error to measure accuracy,
and Spearman's rank-order correlation to measure the ability to correctly rank
populations.

@@

Applied to the simulated dataset of known status, the individual models had
variable success at recovering the mean (status) and slope (trajectory) of
\bbmsy\\ in the last five years. All models exhibited a high degree of scatter
around the 1:1 line of perfect status prediction (Fig. \ref{hexagon}). CMSY
exhibited bimodal predictions of status but had the best rank-order correlation
and accuracy scores (Fig. \ref{performance}a). COM-SIR and SSCOM both correctly
identified a number of low status stocks, but frequently predicted a high
status when status was in fact low (Fig. \ref{hexagon}b, c). mPRM had
relatively poor ability to predict status for the simulated dataset (Fig.
\ref{hexagon}d). There was generally little correlation between true and
predicted trajectory for any of the individual models besides SSCOM (Figs
\ref{scatter-sim-slope}a--d).

@@

The ensemble models were able to improve the predictions of status by
exploiting the best properties of individual models, the covariance between
individual models, and interactions with other covariates. For example, SSCOM
had strong predictive ability when it predicted low \bbmsy\\ (Fig.
\ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were
approximately linearly related to \bbmsy\\ within the low and high prediction
clusters (Fig. \ref{partial-sim}). Ensembles also exploited the covariance
between individual model predictions. For instance, both the linear model and
GBM ensemble suggest that if mPRM and SSCOM predict high status, the true
status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l). The
addition of spectral density covariates helped the ensemble models correctly
predict higher status values (Fig. \ref{hexagon}g, h) by exploiting
interactions between the spectral density predictors and the individual model
predictions (Fig. \ref{partial-2d-sim}). Although the additional covariates
improved ensemble fit, performance of the ensembles was only marginally
degraded by removing these covariates (Fig. \ref{hexagon} vs. Fig.
\ref{hexagon-sim-basic}).

@@

<!--
Some individual models performed nearly as well as the ensembles in one or two performance dimensions: e.g. SSCOM does nearly as well as the ensembles at predicting the slope of \bbmsy\\ (Fig. \ref{performance-sim-slope}). But no individual method is consistently nearly as good as the ensembles. SSCOM does more poorly than the ensembles and other individual models in terms of accuracy, rank-order correlation, and bias at estimating the mean \bbmsy\\ for both the simulated and stock-assessment datasets (Fig. \ref{performance}).
-->

@@

A strength of ensemble models is that they can be tailored or specific response
variables. For example, we built separate ensemble models of mean \bbmsy\\ and
the slope of \bbmsy\\. The same set of model weights or non-linear
relationships need not hold across different response variables. For example,
SSCOM contributed little to the GBM ensemble estimate of status at higher
levels of predicted \bbmsy\\, but contributed strongly to estimates of slope
(TODO FIG). Formally, fitting ensemble models to specific quantities of
interest provides an additional calibration step [@rykiel1996]. Individual
models are typically calibrated (i.e. parameters are estimated) based on a
single response variable; however, this model might then provide a biased or
inaccurate estimate of other quantities of interest. For example, a
data-limited fisheries model might calibrate parameters by maximizing
likelihood with respect to observed and predicted catches but those parameters
may then provide a biased or inaccurate estimate of the mean and slope of
\bbmsy\\. Ensemble methods provide a final calibration step to a quantity of
interest. This ensemble calibration could further include a loss function
tailored to the goals of the model, say placing greater weight on accuracy at
low status levels than higher values.

>>>>

What can we do when multiple ecological models suggest different conclusions about population status? Any single model is unlikely to perform best under all scenarios. Ensemble methods, commonly used in fields such as climate science, provide a powerful tool to leverage information across multiple ecological models. Here we evaluate the potential for ensemble methods to improve estimates of fish population status and trajectory. We fit four data-limited models of population and exploitation dynamics and combine their estimates of the mean and slope of biomass at maximum sustainable yield ($B/B_\mathrm{MSY}$) with four ensemble models: a simple average, a linear model, and two machine-learning methods---random forests and boosted regression trees. We build our ensembles on a simulated dataset of 5760 fish stocks and test our models with cross-validation and a global database of `r ram_stocks_n` stock assessments. We find that ensemble models substantially improve estimates of population status and trajectory. Machine-learning ensembles performed the best: accuracy improved `r mean_sim$mach_vs_ind_mare_fold`%, correlation improved from `r mean_sim$ind_corr_range` to `r mean_sim$mach_corr_range`, and bias (median proportional error) declined from `r mean_sim$ind_mre_range` to `r mean_sim$mach_mre_range`. We found similar improvements when predicting the slope of status. When the simulation-trained ensembles were applied to global fish stocks, the best ensembles still improved all dimensions of performance over the next-best individual model. Ensemble models have potential to improve estimates of population status and trajectory across many taxa beyond fish. 
However, ensemble methods must be carefully tested, formed from a diverse set of accurate models, and built on a dataset sufficiently representative of the populations they are applied to.

@@

We fit four individual data-limited models to estimate $B/B_\mathrm{MSY}$. Three of the
models are mechanistic and based generally on Schaefer biomass dynamics
[@schaefer1954] of the form

@@

* *mPRM* (modified panel regression model) is a modified version of the
  panel-regression model from Costello et al. [@costello2012]. Unlike the other
  models, mPRM is empirical and not mechanistic---it uses the RAM
  Legacy Stock Assessment database to fit a regression model to a series of
  characteristics of the catch time series and stock with stock-assessed
  $B/B_\mathrm{MSY}$ as the response. The model used in this paper is
  modified from the original in that it condenses the life-history categories
  into three categories to match the simulated dataset, removes the maximum
  catch predictor since the absolute catch in the simulated dataset is
  arbitrary, and does not implement the bias correction needed in Costello et
  al. [@costello2012] for deriving aggregate estimates of mean status across
  multiple stocks.

@@

The individual models we seek to combine with ensemble models provide time
series of $B/B_\mathrm{MSY}$. Therefore, we can use ensemble models to estimate multiple
properties of these status time series. Here, we focus on two properties: the
mean and slope of $B/B_\mathrm{MSY}$ in the last 5 years. Together, these quantities
address the recent state and trajectory of status, which are both of management
and conservation interest (e.g. [@mace2008]). To avoid undue influence of the
end points of the time series on the calculated slope, we measured the slope as
the Theil-Sen estimator of median slope [@theil1950].

@@

\noindent 
where $L$ is a generalized loss function, $f$ is some generalized regression
function, $\hat{b}_\mathrm{CMSY}$ is the mean or slope prediction from the CMSY
model etc. (Table \ref{tab:predictors}), $S(0.2)$ and $S(0.05)$ represent the
spectral density of the scaled catch series at frequencies of 0.20 and 0.05, $b$
is the mean or slope value the model is fitted to, and $\theta$ is the set of
parameters for the regression function. For all ensemble models of mean
$B/B_\mathrm{MSY}$---a ratio bounded at zero---we fit the models in log space and
exponentiated the predictions. For the estimates of $B/B_\mathrm{MSY}$ slope, we fit
ensemble models on the natural untransformed scale.

We calculated the average ensemble models as the geometric mean of the
individual model $B/B_\mathrm{MSY}$ estimates and the mean of the individual model
$B/B_\mathrm{MSY}$ slopes. We fit the linear model ensemble with second-order
interactions. We fit two machine learning ensemble models based on regression
trees. Regression trees sequentially determine what value of a predictor best
splits the response data into two 'branches' based on a specified loss
function [@breiman1984]. In random forests, a series of regression trees are
built on a random subset of the data and random subset of the covariates of the
model [@breiman2001]. In generalized boosted models (GBMs), each subsequent
model is fit to the residuals from the previous model; data points that are fit
poorly in a given model are given more weight in the next model [@elith2008].
Random forests and GBMs can provide strong predictive performance and fit
highly non-linear relationships [@elith2008; @hastie2009]. We fit random forest
models with the `randomForest` package [@liaw2002] for \textsf{R} with the
default argument values. We fit boosted regression tree models with the `gbm`
package [@ridgeway2015] for \textsf{R}. Based on cross-validation with the
**caret** \textsf{R} package, we fit GBMs with 2000 trees, an interaction depth
of $6$, a learning rate (shrinkage parameter) of $0.01$, and all other
arguments at their default values.

@@

Predictive performance can be evaluated with metrics that represent a variety
of modelling goals. For continuous response variables such as the mean and
slope of population status, performance metrics often measure some form of
bias, precision, accuracy (a combination of bias and precision), or the ability
to correctly rank or correlate across populations (e.g. [@walther2005]). Here,
we measure proportional error, defined as
$(\hat{b} - b)/b$, where $\hat{b}$ and $b$ represent estimated and 'true'
(or stock-assessed) mean or slope of $B/B_\mathrm{MSY}$. We calculated median proportional
error to measure bias, median absolute proportional error to measure accuracy,
and Spearman's rank-order correlation to measure the ability to correctly rank
populations.

@@

Applied to the simulated dataset of known status, the individual models had
variable success at recovering the mean (status) and slope (trajectory) of
$B/B_\mathrm{MSY}$ in the last five years. All models exhibited a high degree of scatter
around the 1:1 line of perfect status prediction (Fig. \ref{hexagon}). CMSY
exhibited bimodal predictions of status but had the best rank-order correlation
and accuracy scores (Fig. \ref{performance}a). COM-SIR and SSCOM both correctly
identified a number of low status stocks, but frequently predicted a high
status when status was in fact low (Fig. \ref{hexagon}b, c). mPRM had
relatively poor ability to predict status for the simulated dataset (Fig.
\ref{hexagon}d). There was generally little correlation between true and
predicted trajectory for any of the individual models besides SSCOM (Figs
\ref{scatter-sim-slope}a--d).

@@

The ensemble models were able to improve the predictions of status by
exploiting the best properties of individual models, the covariance between
individual models, and interactions with other covariates. For example, SSCOM
had strong predictive ability when it predicted low $B/B_\mathrm{MSY}$ (Fig.
\ref{hexagon}c, Fig. \ref{partial-sim}c) and CMSY predictions were
approximately linearly related to $B/B_\mathrm{MSY}$ within the low and high prediction
clusters (Fig. \ref{partial-sim}). Ensembles also exploited the covariance
between individual model predictions. For instance, both the linear model and
GBM ensemble suggest that if mPRM and SSCOM predict high status, the true
status also tends to be high (Figs \ref{lm-coefs}, \ref{partial-2d-sim}l). The
addition of spectral density covariates helped the ensemble models correctly
predict higher status values (Fig. \ref{hexagon}g, h) by exploiting
interactions between the spectral density predictors and the individual model
predictions (Fig. \ref{partial-2d-sim}). Although the additional covariates
improved ensemble fit, performance of the ensembles was only marginally
degraded by removing these covariates (Fig. \ref{hexagon} vs. Fig.
\ref{hexagon-sim-basic}).

@@

<!--
Some individual models performed nearly as well as the ensembles in one or two performance dimensions: e.g. SSCOM does nearly as well as the ensembles at predicting the slope of $B/B_\mathrm{MSY}$ (Fig. \ref{performance-sim-slope}). But no individual method is consistently nearly as good as the ensembles. SSCOM does more poorly than the ensembles and other individual models in terms of accuracy, rank-order correlation, and bias at estimating the mean $B/B_\mathrm{MSY}$ for both the simulated and stock-assessment datasets (Fig. \ref{performance}).
-->

@@

A strength of ensemble models is that they can be tailored or specific response
variables. For example, we built separate ensemble models of mean $B/B_\mathrm{MSY}$ and
the slope of $B/B_\mathrm{MSY}$. The same set of model weights or non-linear
relationships need not hold across different response variables. For example,
SSCOM contributed little to the GBM ensemble estimate of status at higher
levels of predicted $B/B_\mathrm{MSY}$, but contributed strongly to estimates of slope
(TODO FIG). Formally, fitting ensemble models to specific quantities of
interest provides an additional calibration step [@rykiel1996]. Individual
models are typically calibrated (i.e. parameters are estimated) based on a
single response variable; however, this model might then provide a biased or
inaccurate estimate of other quantities of interest. For example, a
data-limited fisheries model might calibrate parameters by maximizing
likelihood with respect to observed and predicted catches but those parameters
may then provide a biased or inaccurate estimate of the mean and slope of
$B/B_\mathrm{MSY}$. Ensemble methods provide a final calibration step to a quantity of
interest. This ensemble calibration could further include a loss function
tailored to the goals of the model, say placing greater weight on accuracy at
low status levels than higher values.
